Potential of I/O-Aware Workflows in Climate and Weather

### 2020-02-04 paper/paper.tex
Updating

%\documentclass[a4paper, twocolumn]{article}
\documentclass[a4paper]{article}
\newcommand{\papertitle}{Potential of I/O-Aware Workflows in Climate and Weather}

\usepackage[a4paper, margin=2cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage[colorlinks=true,urlcolor=red]{hyperref}
\usepackage{url}
\usepackage{float}
\usepackage{cleveref}
\usepackage{todonotes}
\usepackage{comment}
\usepackage{ulem}
\usepackage{array}
\usepackage{subcaption}

\newcommand{\jk}[1]{\todo[inline]{JK: #1}}
\newcommand{\lr}[1]{\textcolor{blue}{LR: #1}}

\usepackage{multicol}
\setlength{\columnsep}{1cm}

\usepackage{fancyhdr}
\fancyhead{}
\fancyfoot{}
\fancyhead[l]{\papertitle}
\fancyhead[r]{\includegraphics[height=1em]{esiwacelogo_type_grey_left_2c_ext}}
\fancyfoot[r]{\thepage}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}

\usepackage{titling}
\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}\vskip 0.5em}

\graphicspath{{./assets/}}

\title{\papertitle}

\author{
%  Julian M. Kunkel
%  \textit{University of Reading}
%    \and
%  Luciana R. Pedro
%  \textit{University of Reading}
}
\date{\today}

\begin{document}
\maketitle
\thispagestyle{fancy}

\section*{Abstract}
The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and the needs of the end-to-end user.
As climate and weather workflows become increasingly complex and blend beyond data centres while, at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.

The key contributions of this paper are:
1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy;
2) we illustrate architecture and roadmap that allows the seamless integration into existing workflows.
The tools employed here to achieve this extended workflow are Cylc, XIOS, DDN IME, and ESDM.

%Basically, we believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support the programming models.
From the user perspective, we ultimately aim to provide an abstraction to data dependencies.
The proposed changes increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
From the system perspective, the vision aims to derive an execution plan that utilises the available hardware and software infrastructure within a data centre.

\section{Introduction}

@@

Based on their needs, the HPC community has developed a software ecosystem that supports them to execute their large-scale workflows.
While the current advances correspond to a big leap forward, many processes still require experts.
For example, porting a workflow from one system to another requires to adjust runtime parameters of applications and to decide how data is managed.
Since performance is of crucial importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
To obtain the best performance, a specific supercomputer typically requires substantial changes to the workflow to tailor the workflow to the particular machine.
For instance, a data-driven workflow may benefit from using a heterogeneous set of computing and storage technology at the same time.

Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise resource usage in workflows. In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs; 2) users are not aware of the potential of the complex system.

In this paper, we show that knowledge about Input/Output of workflow tasks and overall experimental design helps to
optimise the execution of climate and weather workflows and may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.

%The paper is organised as follows:
%In \Cref{sec:workflows}, we provide an overview of the specification of workflows in climate and weather and the involved software stack.

@@

Future centres are expected to have even more heterogeneity. A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide storage and processing capabilities.
\Cref{fig:heterogeneous} shows such an environment with a focus on computation and storage.
Some of these technologies might be local to specific compute nodes or globally available.
Depending on the need, the storage characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
\jk{Mention BBs, e.g., DDN IME}
Naturally, the variety of tasks executed by a single workflow may benefit from the different storage and computing infrastructure.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{system}
  \caption{Example of an heterogeneous HPC landscape}
  \label{fig:heterogeneous}
\end{figure}

@@

A typical I/O stack for a single parallel application like a climate model is shown in \Cref{fig:layers}.

\begin{minipage}{0.2\textwidth}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{layers-xios}
  \caption{I/O path for an MPI-parallel application}
  \label{fig:layers}
\end{figure}
\end{minipage}
\qquad
\begin{minipage}{0.7\textwidth}
In our example, we assume the application is parallelised for performance reasons and uses MPI to run the application.
It may use XIOS to gather data from individual fields which then uses NetCDF4 to store the data as a file.
XIOS provides a domain-level semantics and ...\jk{TODO}
Under the hood, NetCDF4 used the HDF5 API and file format.
Internally, HDF5 uses MPI again and its data types to specify the nature of the data stored.
Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in the local file system as block devices on storage media such as SSDs and HDDs.

@@

Within ESiWACE, we are developing ESDM, ... \jk{TODO}
\end{minipage}

@@

The appropriate software stack involved in executing a workflow is depicted in \Cref{fig:stages}.
As a preliminary, the user had specified a Cylc suite representing the workflow and provided scripts files for the individual tasks to execute; now the user enacts Cylc to start the workflow.
In the following, each stage of the execution is described further.

\begin{figure}[H]
  \centering
  \includegraphics[scale=1.4]{stages}
  \caption{Software stack and stages of execution}
  \label{fig:stages}
\end{figure}

\begin{enumerate}
  \item \textbf{Cylc} Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
  Cylc monitors the progress of the workflow.
  Once a task can be executed (dependencies are fulfilled), Cylc generates a Slurm script on the fly with the required metadata for Slurm that will invoke the user-defined script for that task and runs Slurm to queue the job.

  \item \textbf{Slurm} Slurm schedules the execution of the job considering the scheduling policy of the data centre.
  Once the job is ready to be dispatched, i.e., resources are available, and the job priority is the highest, it is started on the supercomputer.

  \item \textbf{Job} The user-provided script is executed on a single node assigned to the job.
  The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).

  \item \textbf{Script} The shell-script executes the sequence of commands, including applications that are parallel programs.
  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application which, ultimately, defines the storage location to be used.
  A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.

  \item \textbf{Application} The user-specified parallel application runs taking the filenames as defined in a configuration file or as arguments.
  The application may use NetCDF or XIOS to perform the I/O.
\end{enumerate}

Consider the Cylc suite configuration for a toy monthly cycling workflow in \Cref{fig:cylc} \cite{8675433}.
In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
This process is repeated in the model to simulate month and years into the future.
Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
A user has specified a detailed workflow in a Cylc suite file (\texttt{suite.rc}).
In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\columnwidth]{cylc1}
  \caption{Example workflow}
  \label{fig:cylc}
\end{figure}

This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
Decisions about storage must take into consideration the architecture in which the workflow will run.
The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.

@@

The scripts for the tasks in Cylc define how the data is stored on the available storage systems.
Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.

What happens in many current workflows is that they use the only \textbf{work} and \textbf{scratch}.
When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
If the \textbf{scratch} filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until it is finished.
If the I/O planning is made a priori, this situation can be avoided.
In addition to the file systems described, many supercomputers also have available for the user \textbf{scratch2} file systems (sometimes local and global).
Therefore, in a real scenario, if the architecture of the machine is known, we can better utilise its resources.

\Cref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
As shown in \Cref{fig:lifecycle1}, the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
In the end, this dataset might be an intermediate product that can then be deleted.
Alternatively (\Cref{fig:lifecycle2}, it could be stored on the \textbf{scratch} file system immediately and accessed there.
However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
The last scenario (\Cref{fig:lifecycle3}) presents the case where data is created on \textbf{work} by a task. Still, then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
However, it would require that the subsequent tasks are placed on the same machine where data is now stored.
The scientist responsible for the workflow must optimise the available resources intuitively.
These are just examples, and currently, the scientists perform such kind of optimisations manually.

\begin{figure}[H]
  \centering
  \begin{subfigure}{.3\textwidth}
  \includegraphics[width=0.9\columnwidth]{lifecycle-1}
  \caption{\label{fig:lifecycle1} Local and work file systems}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
  \includegraphics[width=0.9\columnwidth]{lifecycle-2}
  \caption{\label{fig:lifecycle2} Scratch file system only}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
  \includegraphics[width=0.9\columnwidth]{lifecycle-3}
  \caption{\label{fig:lifecycle3} Local and work file systems}
  \end{subfigure}
  \caption{Alternative life cycles for mapping a dataset to storage and the operations: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
  \label{fig:lifecycle}
\end{figure}

@@

\section{Vision for I/O-Aware Workflows}

@@

Our vision requires two additional information: firstly, the user must augment the workflow description with information about I/O requirements and explicitly annotate dependencies to data sets; secondly, the data centre (or expert user) must provide a description of the storage architecture and characteristics for each storage system.

There are many approaches to implement the technology for this vision. It is apparent that changes need to be made to several software components to realise this vision.
In the next section, we will discuss the specific design for our transitional roadmap for climate and weather.

@@

So far, Cylc allowed specifying tasks and the dependencies among them.
We aim to enrich this information with characteristics for input, and output, i.e., the datasets used.
An example workflow containing input data sets and (intermediate) products is illustrated in \Cref{fig:workflow}.
Nodes represent tasks or data; arrows indicate dependencies among between them.
In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
Most of the workflow can run automatically, except for the manual quality control of the products and the final data usage of Product\,3.
This last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
In this approach, each task is annotated with the required inputs and the generated products.
Each data product is annotated with metadata such as data life cycle information, e.g., the value of data and how long should it be kept.

%It was correctly placed here

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\columnwidth]{workflow}
  \caption{Example of a high-level workflow with tasks and data dependencies}
  \label{fig:workflow}
\end{figure}

@@

The system information shall comprise of all available storage systems, the system topology, and the characteristics of each of its required components.
This may include simplified or complex models of the components to approximate performance for specific I/O patterns allowing to predict performance behaviour.
It is expected that the data centre (or expert users) can create such a configuration file, for instance, by using vendor-provided information or by executing benchmarks.

@@

With this separation of concerns, we can abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
In this paper, we focus on utilising a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.
Every time parameters are manually tuned right now there is a room from improvement by using automatised decision making.
And Machine Learning has been proven effective in replacing human-beings and adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).

Optimisations we discuss in this paper are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
Firstly, the I/O scheduler could make the placement decision for the individual data sets and manage their relocation during the life cycle, as discussed in \Cref{sec:datamanagement}.

@@

\Cref{fig:cycle4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
Assume you have, for instance, four cycles of the model running with all the necessary post-processing steps.
Typically, the user has written the scripts for each task by defining a filename using the same prefix that is then placed at the same storage type every time.
That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
The user-specific storage locations are a viable option to run a workflow.
The user might even have analysed various scenarios such as depicted in
\Cref{sec:datamanagement}.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\columnwidth]{cycle4}
  \caption{Four stages workflow example}
  \label{fig:cycle4}
\end{figure}

@@

Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user, usually considering only options for feasibility.

\begin{itemize}
\item Abstraction: The user does not have to worry about the architecture that the workflow will run, removing the specialist from the decision-making process.

\item Optimisation: The workflow will be optimised for the specific available system characteristics considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
By knowing the workflow behind the data creation, the system could trade data security with potential recomputation opportunities to optimise the cost-efficiency.

\item Performance-portability: The user has to specify the I/O requirements only once for the tasks of the workflow.
The workflow can now be run on any system without user intervention.
Even more, if the system characteristics change, e.g., it gets upgraded, and an additional storage tier becomes available, or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.

@@

\item Adapts the resulting workflow, individual scripts and application executions to consider the automatic I/O schedule.

@@

@@

\Cref{fig:cycle1} introduces one cycle of the suite in \Cref{fig:cylc} with a simplified notation. The arrows represent the tasks dependencies, but there is no information about the data that are generated. Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependencies?
Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create to reproduce the workflow.

\begin{comment}

\begin{figure}[H]
\centering
\begin{tabular}{ccc}
\includegraphics[scale=0.6]{cycle1} & \qquad &
\includegraphics[scale=0.6]{cycle-io-dep} \\
(a) Original Cycle & & (b) Cycle including datasets and dependencies
\end{tabular}
\caption{ \label{fig:cycle} Simplified Cylc workflow for the first cycle}
\end{figure}

\end{comment}

\begin{figure}[H]
  \begin{subfigure}{0.45\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{cycle1}
    \caption{Original Cycle}
    \label{fig:cycle1}
  \end{subfigure}
  \qquad
  \begin{subfigure}{0.45\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{cycle-io-dep}
    \caption{Cycle including datasets and dependencies}
    \label{fig:cycle-io-dep}
  \end{subfigure}
  \caption{Simplified Cylc workflow for the first cycle}
\end{figure}

The scenario presented in \Cref{fig:cycle1} shows the same cycle of \Cref{fig:cycle-io-dep}, but  extended information about datasets is added.
In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool. The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.

@@

The user provides information about datasets required for input and the generated output in a separate file similarly to Cylc's suite file.
This information contains dataset information for each Cylc task:
\begin{itemize}
  \item name: the name of the field generated/required, this may include a pattern such as supported by Cylc
  \item size: an estimate of the file size
  \item lifetime: how long the data must be retained on storage (if at all)
  \item type: may specify the type of the data, i.e., checkpoint, diagnostics, temporary
\end{itemize}
% Expressions will support generic variables such as the number of nodes a job runs...

@@

The ESDM Scheduler will make the schedule. The ESDM Scheduler can be used in different steps of the process. ESDM has already an interface with NetCDF. Assuming the user can provide extra information in advance, ESDM can interact in the process before the user gives the data to the Cylc tool, using the information to construct an optimised workflow. Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.

>>>>

%% This is an article based on superfri.cls class file of
%% ``Supercomputing Frontiers and Innovations. An International Journal''
%% http://superfri.org/.

\documentclass{superfri}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}

\usepackage{todonotes}
\usepackage{cleveref}
\usepackage{subcaption}
\usepackage{comment}

\newcommand{\jk}[1]{\todo[inline]{JK: #1}}
\newcommand{\lr}[1]{\textcolor{blue}{LR: #1}}
\pagestyle{plain}

% ------------
\bibliographystyle{superfri}

\begin{document}

\author{}
%\author{Julian M. Kunkel\footnote{\label{uread}University of Reading, Reading, United Kingdom}\orcidID{0000-0002-6915-1179} \and Luciana R. Pedro\footnoteref{uread}\orcidID{0000-0001-8365-6264}
%}

\title{Potential of I/O-Aware Workflows in Climate and Weather}

\maketitle{}

\begin{abstract}%
The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and the needs of the end-to-end user.
As climate and weather workflows become increasingly complex and blend beyond data centres while, at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.

The key contributions of this paper are:
1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy, and 2) we illustrate architecture and roadmap that allows the seamless integration into current climate and weather workflows.
The tools employed here to achieve this extended workflow are Cylc, XIOS, DDN IME, and ESDM.

We believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support the programming models.
The changes proposed in this work increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.

\keywords{workflow, climate, weather, heterogeneous storage, data-driven}
\end{abstract}

% -----------------------------------------------------------------------
\section*{Introduction}
\label{sec:intro}

@@

Based on their needs, the HPC community has developed a software ecosystem that supports them to execute their large-scale workflows.
While the current advances correspond to a big leap forward, many processes still require experts.
For example, porting a workflow from one system to another still requires adjusting runtime parameters of applications and deciding on how data is managed.
Since performance is of crucial importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
To obtain the best performance, a specific supercomputer typically requires substantial changes to the workflow to tailor the workflow to the particular machine.
For instance, a data-driven workflow may benefit from using a heterogeneous set of computing and storage technology at the same time.

Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise resource usage in workflows.
In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs, and 2) users are not aware of the potential of the complex system.

In this paper, we illustrate how knowing the Input/Output (I/O) characteristics of workflow tasks and overall experimental design help to optimise the execution of climate and weather workflows. This enlightenment may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected any longer.
Our approach intends to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
Once we have an automated decision about where the job will run and how the storage will be managed, climate and weather scientists can then run their workflows in any machine efficiently without prior knowledge about its architecture.

@@

Future centres are expected to have even more heterogeneity. A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide storage and processing capabilities.
\figref{pic/system} shows such an environment with a focus on computation and storage.
Some of these technologies might be local to specific compute nodes or globally available.
Depending on the need, the storage characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
\jk{Mention BBs, e.g., DDN IME}
Naturally, the variety of tasks executed by a single workflow may benefit from the different storage and computing infrastructure.

\fig{width=0.6\columnwidth}{pic/system}{Example of an heterogeneous HPC landscape}

@@

A typical I/O stack for a single parallel application like a climate model is shown in \figref{pic/layers-xios}.

\fig{width=0.2\textwidth}{pic/layers-xios}{I/O path for an MPI-parallel application}

In our example, we assume the application is parallelised for performance reasons and uses MPI to run the application.
It may use XIOS to gather data from individual fields which then uses NetCDF4 to store the data as a file.
XIOS provides a domain-level semantics and ...\jk{TODO}
Under the hood, NetCDF4 used the HDF5 API and file format.
Internally, HDF5 uses MPI again and its data types to specify the nature of the data stored.
Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in the local file system as block devices on storage media such as SSDs and HDDs.

@@

Within ESiWACE, we are developing ESDM, ... \jk{TODO}

@@

The appropriate software stack involved in executing a general workflow is depicted in \figref{pic/stages}.
As a preliminary, the user has to specify a configuration representing the workflow and provide scripts for the individual tasks to be executed. After that, the user enacts a workflow engine to start the workflow.
In the following, each stage of the execution is described further.

\fig{width=0.8\textwidth}{pic/stages}{Software stack and stages of execution}

  \item \textbf{Scientist} specifies a configuration representing the workflow and provides scripts for each task.

  \item \textbf{Workflow Engine} parses the workflow configuration file, generate tasks dependencies, defines a schedule for the execution, and monitors the progress of the workflow.
  Once a task can be executed (dependencies are fulfilled), the workflow engine generates a Slurm script on the fly with the required metadata that will load the user-defined script for each task. Then, Slurm runs to queue the job.

  \item \textbf{Slurm} \cite{Jette02slurm:simple} schedules the execution of the job, considering the scheduling policy of the data centre.
  Once the job is ready to be dispatched, i.e., resources are available, and the job priority is the highest, it is started on the supercomputer.

  \item \textbf{Job} runs the user-provided script on one single node assigned by Slurm.
  The script has set environment variables containing information about the execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).

  \item \textbf{Script} typically calls commands to create a filename with a template provided by the user and feeds this information to the application which, ultimately, defines the storage location.
  A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.

  \item \textbf{Application} runs taking the generated filenames that the script has set.
  The user-specified parallel application may use XIOS \cite{xios} or NetCDF \cite{Jenter92netcdf:a} to perform the I/O.

@@

Usually, the scripts for tasks in climate and weather workflows define how the data is stored on the available storage systems.
What happens in many current workflows is that they ignore the benefits of using multiple file systems concurrently or exploiting the data locality between tasks by co-locating the tasks.
On top of that, in the current state-of-the-art, scientists have to optimise the available storage resources intuitively and insert the information about this decision-making process manually.
Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user, usually considering only options for feasibility. Machine Learning has been proven effective in replacing human-beings, and it may adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).

Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
What happens in many current workflows is that they use the only \textbf{work} and \textbf{scratch}.
When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
If the \textbf{scratch} filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until it is finished.
If the I/O planning is made a priori, this situation can be avoided.
In addition to the file systems described, many supercomputers also have available for the user \textbf{scratch2} file systems (sometimes local and global).
Therefore, in a real scenario, if the architecture of the machine is known, we can better utilise its resources.

\figref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
As shown in \figref{fig:lifecycle1}, the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
In the end, this dataset might be an intermediate product that can then be deleted.
Alternatively (\figref{fig:lifecycle2}, it could be stored on the \textbf{scratch} file system immediately and accessed there.
However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
The last scenario (\figref{fig:lifecycle3}) presents the case where data is created on \textbf{work} by a task. Still, then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
However, it would require that the subsequent tasks are placed on the same machine where data is now stored.
The scientist responsible for the workflow must optimise the available resources intuitively.
These are just examples, and currently, the scientists perform such kind of optimisations manually.

\begin{figure}[b]
    \begin{minipage}{.33\linewidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{pic/lifecycle-1}
        \subcaption{Local and work file systems}\label{fig:lifecycle1}
    \end{minipage}
    \begin{minipage}{.33\linewidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{pic/lifecycle-2}
        \subcaption{Scratch file system only}\label{fig:lifecycle2}
    \end{minipage}
    \begin{minipage}{.33\linewidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{pic/lifecycle-3}
        \subcaption{Local and work file systems}\label{fig:lifecycle3}
    \end{minipage}
    \vspace{5pt}
    \caption{Alternative life cycles for mapping a dataset to storage and the operations: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
    \label{fig:lifecycle}
\end{figure}

@@

\section{Vision for I/O-Aware Workflows}
\label{sec:vision}

There are several approaches to implement the technology for this vision, and changes need to be made to software components to realise it.
In the next sections, we will discuss a specific design for our transitional roadmap considering climate and weather workflows and the tools scientists from this field already use in their routine research.

@@

Our vision for I/O-aware workflows requires two additional information. Firstly, the user must augment the workflow description with information about I/O requirements and explicitly annotate dependencies to data sets. Secondly, the data centre (or expert user) must provide details about the storage architecture.

@@

In general, climate and weather workflows allow specifying tasks and the dependencies among them.
We aim to enable enriching this information with characteristics for input, and output, i.e., the datasets.
An example workflow containing input datasets and (intermediate) products is illustrated in \figref{pic/workflow}.
Nodes represent tasks or data, and arrows indicate the dependencies.
In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
Most of the workflow can run automatically, except for the manual quality control of the products and the final data usage of Product\,3.
This last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
In the approach proposed in this work, each task is annotated with the required input datasets and the generated products with their characteristics.
Each data product is annotated with metadata such as data life cycle information, e.g., the value of data and how long should it be kept.

@@

\fig{width=0.4\columnwidth}{pic/workflow}{Example of a high-level workflow with tasks and data dependencies}

@@

While many optimisations are possible once the abstraction is in place, here we focus on utilising a heterogeneous storage environment. The optimisations we discuss are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.

The system information shall comprise of all available storage systems, the system topology, and details of each of its required components.
This may include simplified or complex models of the components to approximate performance for specific I/O patterns allowing to predict performance behaviour.
It is expected that the data centre (or expert users) can create such a configuration file, e.g., by using vendor-provided information or by executing benchmarks.
With this information, the I/O scheduler can make the placement and migration decisions for individual datasets and manage their relocation during their life cycle, as discussed in \figref{sec:datamanagement}.\lr{Check ref.}
This separation of concerns allows us to abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.

@@

In this paper, we focus on utilising a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.

@@

\figref{pic/cycle4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
Assume you have, for instance, five cycles of the model running with all the necessary post-processing steps.
Typically, the user has written the scripts for each task by defining a filename using the same prefix that is then placed at the same storage type every time.
That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
The user-specific storage locations are a viable option to run a workflow.
The user might even have analysed various scenarios such as depicted in
Section \ref{sec:datamanagement}.

@@

\fig{width=0.8\columnwidth}{pic/cycle4}{Four stages workflow example}

@@

The benefits of the proposed vision are:

\begin{itemize}

\item Abstraction: The user does not have to worry about the architecture that the workflow will run, removing the specialist from the decision-making in the storage process.

\item Optimisation: The workflow will be optimised specifically for the available system infrastructure considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
By knowing the workflow behind the data creation, the system could even trade data security with potential recomputing opportunities to optimise the cost-efficiency.

\lr{Introduce the picture as an example?}

\item Performance-portability: The user has to specify the I/O requirements only once for the tasks of a specific workflow, and the workflow can now run on any system without user intervention.
Even more, if the system characteristics change, e.g., it gets upgraded, an additional storage tier becomes available, or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.

@@

\item Adapts the resulting workflow, individual scripts and application executions to consider the automatic I/O scheduler.

@@

Consider the Cylc workflow for a toy monthly cycling workflow in \figref{pic/cylc1} \cite{8675433}.
In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
This process is repeated in the model to simulate month and years into the future.
Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
A user has specified a detailed workflow in a Cylc configuration file (\texttt{conf.rc}).
In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.

\fig{width=0.9\columnwidth}{pic/cylc1}{Example workflow}

This workflow presents the dependencies among the tasks, but it is missing the information about I/O and storage.
Decisions about storage must take into consideration the architecture in which the workflow will run.
The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.

@@

\figref{fig:cycle1} introduces one cycle of the workflow in \figref{pic/cylc1} with a simplified notation. The arrows represent the tasks dependencies, but there is no information about the data that are generated. Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependencies?
Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create to reproduce the workflow.

\begin{figure}[b]
    \begin{minipage}{.45\linewidth}
        \centering
        \includegraphics[width=0.9\linewidth]{pic/cycle1}
        \subcaption{Including tasks and dependencies}\label{fig:cycle1}
    \end{minipage}
    \begin{minipage}{.45\linewidth}
        \centering
        \includegraphics[width=0.9\linewidth]{pic/cycle-io-dep}
        \subcaption{Including tasks, datasets and dependencies}\label{fig:cycle-io-dep}
    \end{minipage}
    \vspace{5pt}
    \caption{Example of a high-level workflow}
    \label{fig:cycle}
\end{figure}

The scenario presented in \figref{fig:cycle1} shows the same cycle of \figref{fig:cycle-io-dep}, but  extended information about datasets is added.
In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool. The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.

@@

The user provides information about datasets required for input and the generated output in a separate file similarly to Cylc's workflow configuration file.
This information contains dataset information for each Cylc task:

  \item[Name] The name of the field generated/required, which may include a pattern such as supported by Cylc.
  \item[Size] An estimate of the file size.
  \item[Lifetime] How long the data must be retained on storage (if at all).
  \item[Type] It may specify the type of data, i.e., checkpoint, diagnostics, temporary.

% Expressions will support generic variables such as the number of nodes a job runs...

@@

ESDM has already an interface with NetCDF. The ESDM Scheduler will make the schedule, and it can be used in different steps of the process. Assuming the user can provide extra information in advance, ESDM can interact in the process before the user gives the data to the Cylc tool, using the information to construct an optimised workflow. Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
